---
title: "Application of Central Limit Theorem to the exponential distribution"
author: "Alex Gaggin"
output:
  pdf_document:
    fig_height: 3.6
  html_document: default
---
\fontfamily{cmr}
\fontsize{10}{16}
\selectfont

## Overview

Central Limit Theorem (CLT) states that if we take sufficiently large number
of samples of some distribution,
then means of these samples will be distributed in a different way - they
will follow normal distribution, and their average (mean of means) will
be centered on theoretical mean of the original distribution.
The following simulation demonstrates how this applies to exponential
distribution. Exponential distribution is described by a rate parameter
lambda, and then its theoretical mean and standard deviation will both be
1/lambda.

## Simulations
```{r include=FALSE,eval=FALSE}
Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.

We can use rexp() function in R to simulate random observations, following
the exponential distribution. Following code creates list "dat" of 1000 samples
of 40 observations each, and then takes means of these samples.
```
We use rexp() function in R to simulate random observations.

Let's set simulation parameters and calculate theoretical values for them.
```{r}
lambda         <- 0.2
mu             <- 1/lambda                # Theoretical mean
sigma          <- 1/lambda                # Theoretical standard deviation
variance       <- sigma^2                 # Theoretical variance
sample.size    <- 40                      # Sample size
numsim         <- 1000                    # Number of simulations
SEM            <- sigma/sqrt(sample.size) # Theoretical standard error of the mean
var.of.mean    <- SEM^2                   # Theoretical variance of the mean
```
Let's generate set number of samples and calculate properties
of the distrinution of their means.
```{r}
dat <- list(); set.seed(1)
for(i in 1:numsim) dat[[i]] <- rexp(sample.size, lambda)
means <- sapply(dat, mean)
sample.mean <- mean(means); sample.var <- var(means); sample.sd <- sd(means)
```

## Sample Mean versus Theoretical Mean
```{r include=FALSE,eval=FALSE}
Include figures with titles. In the figures, highlight the means you are
comparing. Include text that explains the figures and what is shown on them,
and provides appropriate numbers.
```
The CLT suggests that mean of means of sufficiently large number of samples
should converge to theoretical mean of the original distribution. 
```{r}
cat(paste("Mean is",sample.mean,"in the simulation and", mu, "in the theory."))
```
Sample and Theoretical means are almost indistinguishable in the histogram
of the sample means (see the R code
for the plot in the Appendix 3).

```{r, echo=FALSE}
breaks <- 20
hist(means, breaks=breaks,
     main="Fig 1. Sample Mean consistently estimates Theoretical Mean",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
abline(v=sample.mean, lwd=2, lty=2, col="blue")
abline(v=mu, lwd=2, col="red")
legend("topright", c(paste0("Sample Mean = ", round(sample.mean, 2)),
                     paste0("Theoretical Mean = ", mu)),
       lty=c(2,1), lwd=c(2,2),col=c("blue","red"))
```

## Sample Variance versus Theoretical Variance
```{r include=FALSE,eval=FALSE}
Include figures (output from R) with titles. Highlight the variances you are
comparing. Include text that explains your understanding of the differences
of the variances.
```
Variance of sample means for sufficiently large number of samples should
converge to squared theoretical Standard Error of the Mean. Their distribution is
more focused
than the original distribution, because the more samples we collect, the less error
and less variability is in the collected data. At infinite number of samples,
n would be infinity, SEM and theoretical variance of the sample means
distribution would be zero, and sample mean will be exactly equal to theoretical
mean of the original distribution. But we have limited number of samples, so
let's see how close sample and theoretical variances of the means are.
```{r}
cat(paste("Variance is",sample.var,"in the simulation and", var.of.mean, "in the theory."))
```
For the chart, let's plot range of two standard deviation, because the
variance is in squared units and can't be visualized directly (see the R code
for the plot in the Appendix 3).

```{r, echo=FALSE}
hist(means, breaks=breaks,
     main="Fig 2. Sample Variance estimates Theoretical Variance
(plotted as standard deviation ranges)",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
abline(v=mu - SEM, lwd=1, lty=2, col="red")
abline(v=mu + SEM, lwd=1, lty=2, col="red")
abline(v=sample.mean - sample.sd, lwd=2, lty=3, col="blue")
abline(v=sample.mean + sample.sd, lwd=2, lty=3, col="blue")
legend("topright", c(paste0("Range of two\nSample Standard\nDeviations = ",
                            round(2 * sample.sd, 2), "\n"),
                     paste0("Range of two\ntheoretical SEM = ",
                            round(2 * SEM, 2))),
       lty=c(3,2), lwd=c(2,1),col=c("blue","red"), bty="n", cex=0.8)
```

## Distribution
```{r include=FALSE,eval=FALSE}
Via figures and text, explain how one can tell the distribution is approximately
normal.
```
Histogram's shape is bell-like, let's draw a normal distribution's bell (with
the theoretical mu and SEM) on top of it to illustrate this (see the R code
for the plot in the Appendix 3).

```{r, echo=FALSE}
hist(means, breaks=breaks, main="Fig 3. Distribution of sample means
is approximately normal",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
curve(192*dnorm(x, mean=mu, sd=SEM), 
      col="darkblue", lwd=2, add=TRUE, yaxt="n")
```

# Appendices

## Appendix 1. Shape of exponential distribution
This is the shape of exponential distribution. We can look at it by plotting
its density function.
```{r}
curve(dexp(x),-1,7)
```

## Appendix 2. Look at a random sample

If we take a random sample and see it, we should see something similar to the
previous plot of the distribution function for the exponential distribution.
And its mean and standard deviation should be close to 5 both.
```{r}
set.seed(1)
random.sample <- sample(1:numsim,1)
hist(dat[[random.sample]],
     main=paste0("Fig 4. Distribution in a sample, sample N", random.sample),
     xlab="Exponentials")

(random.mean <- mean(dat[[random.sample]]))
(random.sd <- sd(dat[[random.sample]]))
```
## Appendix 3. R code for plots
R code for Fig 1.
```{r, eval=FALSE}
breaks <- 20
hist(means, breaks=breaks,
     main="Fig 1. Sample Mean is a consistent estimator of the Theoretical Mean",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
abline(v=sample.mean, lwd=2, lty=2, col="blue")
abline(v=mu, lwd=2, col="red")
legend("topright", c(paste0("Sample Mean = ", round(sample.mean, 2)),
                     paste0("Theoretical Mean = ", mu)),
       lty=c(2,1), lwd=c(2,2),col=c("blue","red"))
```
R code for Fig 2.
```{r, eval=FALSE}
hist(means, breaks=breaks,
     main="Fig 2. Sample Variance estimates Theoretical Variance
(plotted as standard deviation ranges)",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
abline(v=mu - SEM, lwd=1, lty=2, col="red")
abline(v=mu + SEM, lwd=1, lty=2, col="red")
abline(v=sample.mean - sample.sd, lwd=2, lty=3, col="blue")
abline(v=sample.mean + sample.sd, lwd=2, lty=3, col="blue")
legend("topright", c(paste0("Range of two\nSample Standard\nDeviations = ",
                            round(2 * sample.sd, 2), "\n"),
                     paste0("Range of two\ntheoretical SEM = ",
                            round(2 * SEM, 2))),
       lty=c(3,2), lwd=c(2,1),col=c("blue","red"), bty="n", cex=0.8)
```
R code for Fig 3.
```{r, eval=FALSE}
hist(means, breaks=breaks, main="Fig 3. Distribution of sample means
is approximately normal",
     xlab="Means of samples from exponential distribution",
     ylab="Number of samples with means in each bin")
curve(192*dnorm(x, mean=mu, sd=SEM), 
      col="darkblue", lwd=2, add=TRUE, yaxt="n")
```
Because normal distribution line at the figure above is drawn on the same
vertical scale as the histogram, it had to be vertically scaled. Let's use
another simulation to determine the scale, assuming that hist()'s plot function
algorithm isn't transparent to us, and so we don't know a way to calculate this
theoretically. In order to make it more symmetrical, number of observations
should be largely increased.
```{r}
larger <- 1000
round(max(hist(rnorm(numsim * larger), breaks=breaks, plot=FALSE)$counts)/larger)
```
