---
title: "Research of confidence intervals in effect
of Vitamin C supplement on guinea pig tooth growth"
author: "Alex Gaggin"
output:
  pdf_document:
    fig_height: 3.6
  html_document: default
---
\fontfamily{cmr}
\fontsize{10}{16}
\selectfont
```{r include=FALSE}
library(dplyr)
library(datasets)
library(ggplot2)
```

## Overview
Confidence intervals in ToothGrowth dataset show that Vitamin C supplement
increases length of guinea pigs' tooth, and orange juice should be preferrable
way of delivering supplement at lower doses.

## Basic exploratory data analyses

```{r}
data(ToothGrowth)
# ?ToothGrowth
head(ToothGrowth)
summary(ToothGrowth)
str(ToothGrowth)
print(ggplot(data=ToothGrowth, aes(x=dose, y=len, shape=supp)) +
             xlab("Dose") +
             ylab("Tooth length") +
      ggtitle("ToothGrowth data: length vs dose") +              
      geom_point(size=6, alpha=0.5, fill="red") + #guides(shape=FALSE) +
        scale_shape_manual(values=c(22, 24),
                           name="Supplement type",
                       breaks=c("OJ", "VC"),
                       labels=c("Orange Juice", "Ascorbic Acid"))  
              )
```
ToothGrowth dataset describes response in the length of teeth of 10 guinea pigs
treated with Vitamin C. There are six groups produced for three dose levels
(0.5, 1 and 2 mg) and two supplement delivery methods (orange juice and ascorbic
acid).  

# Confidence intervals

Intuitive interpretation of the data on the chart is that supplement does
affects tooth length, but delivery type efficiency depends on dose. Let's
verify by calculating confidence intervals.

It's not completely clear if 10 guinea pigs described
in the help page for ToothGrowth dataset are the same animals in each of groups,
or different ones. Even if they are the same, there's no clear indication in the
dataset of particular rodents, so we are going to treat them as if they were
different, and data points aren't paired. If we will still see that changes
are signigicant, then it will be true even if pigs were the same and data points
paired, because paired data is less variable, and paired confidence intervals are
narrower.

## Length per dose
We calculate confidence intervals for tooth length change for four pair of groups:
dose change from 0.5 to 1 and dose  change from 1 to 2 - for each type of
supplement.
Notice that t.test()'s default value for paired is FALSE, and this is what we want.
We also presume that variability in groups isn't equal and
this is also the default setting for the t test.
```{r}
CheckIntervalByDose <- function(before,after,supp) {
      if(missing(supp)) suppC <- TRUE else suppC <- ToothGrowth$supp == supp
      t.test(ToothGrowth$len[ToothGrowth$dose == after & suppC],
       ToothGrowth$len[ToothGrowth$dose == before & suppC]
       )$conf.int %>% cat
}
CheckIntervalByDose(0.5, 1, "OJ")
CheckIntervalByDose(1,   2, "OJ")
CheckIntervalByDose(0.5, 1, "VC")
CheckIntervalByDose(1,   2, "VC")
```
We see that the confidence intervals lie entirely above zero in every case,
so it's clear that increasing the dosage of the supplement does indeed increase
tooth length in guinea pigs with 95% confidence level.

## Length per supplement
Does supplement delivery affect tooth length?

```{r}
print(ggplot(ToothGrowth, aes(x=supp, y=len, fill=supp)) + geom_boxplot() +
        ylab("Tooth length") + guides(fill=FALSE) +xlab("Delivery type") +
        facet_wrap(~dose))
```
It looks like for lower doses orange juice is more efficient delivery mechanism,
while for 2mg dose it doesn't matter anymore. Let's check these assumptions
by calculating confidence intervals, assuming data isn't paired and variance
can be different for each group.
```{r}
CheckIntervalBySupp <- function(dose) {
      t.test(ToothGrowth$len[ToothGrowth$dose == dose & ToothGrowth$supp == "OJ"],
       ToothGrowth$len[ToothGrowth$dose == dose & ToothGrowth$supp == "VC"]
       )$conf.int %>% cat
}
CheckIntervalBySupp(0.5)
CheckIntervalBySupp(1)
CheckIntervalBySupp(2)
```
Indeed, we see that it's only at higher dosage of 2mg we can't say if vitamin C
delivery method makes a difference. But for 0.5mg and 1mg orange juice is
better, as confidence intervals in these cases lie entirely above zero.
Efficiency doesn't seem to be dependant on the dose in a linear way though,
as confidence for 0.5mg and 1mg aren't very different.

# Result
Assuming data is normally distributed, and using t distribution for safer
approximation for small given number of observations, the confidence intervals
show that

* change in teeth length is statistically significant for each
level of dose increase,
* orange juice is more effective delivery mechanism for vitamic c for lower
doses of the supplement.

# Appendices

### Appendix 1. Effect of ignoring data structure aspects
It probably wouldn't make much sense for a real research to lose
information we have about the experiment,
but just for the training purposes, let's see if the result changes when we do
that.

First, let's ignore supplement delivery type when checking length dependance on
the dose.
```{r}
CheckIntervalByDose(0.5, 1)
CheckIntervalByDose(1,   2)
```
Intervals are still entirely above zero, and the distances aren't dramatically
different than before - it's probably because increased variability was
compensated by having twice as much data points, and having more points
increases confidence.

Now, let's try to see delivery type effect without separation by dose.
```{r}
print(ggplot(ToothGrowth, aes(x=supp, y=len, fill=supp)) + geom_boxplot() +
        ylab("Tooth length") + guides(fill=FALSE) +xlab("Delivery type"))
```
The plot looks like orange juice is better, but is it?
```{r}
t.test(ToothGrowth$len[ToothGrowth$supp == "OJ"],
       ToothGrowth$len[ToothGrowth$supp == "VC"]
       )$conf.int %>% cat
```

We see that if variance isn't the same for each group, and rodents were
different in each group, then we can't say that in general orange juice is
more effective, because our confidence interval touches zero.

### Appendix 2. Effect of presumed equal variance

By the way, what if we are to presume equal variance here, will this assumption
tip the scales?
```{r}
t.test(ToothGrowth$len[ToothGrowth$supp == "OJ"],
       ToothGrowth$len[ToothGrowth$supp == "VC"],
       var.equal=TRUE
       )$conf.int %>% cat
```
Well, interval became a bit narrower but not narrow enough to miss zero. So
in general, orange juice isn't more effective.