---
title: "Research of confidence intervals in effect
of Vitamin C supplement on guinea pig tooth growth"
author: "Alex Gaggin"
output:
  pdf_document:
    fig_height: 4.5
  html_document: default
---
\fontfamily{cmr}
\fontsize{10}{16}
\selectfont

```{r include=FALSE}
library(dplyr)
library(datasets)
library(ggplot2)
```

## Overview

Confidence intervals in ToothGrowth dataset show that Vitamin C supplement
does indeed increase length of guinea pigs' tooth.

## Basic exploratory data analyses


```{r}
data(ToothGrowth)
# ?ToothGrowth
head(ToothGrowth)
summary(ToothGrowth)
str(ToothGrowth)
print(ggplot(data=ToothGrowth, aes(x=dose, y=len, shape=supp)) +
             xlab("Dose") +
             ylab("Tooth length") +
      ggtitle("ToothGrowth data: length vs dose") +              
      geom_point(size=6, alpha=0.5, fill="red") + #guides(shape=FALSE) +
        scale_shape_manual(values=c(22, 24),
                           name="Supplement type",
                       breaks=c("OJ", "VC"),
                       labels=c("Orange Juice", "Ascorbic Acid"))  
              )
```

ToothGrowth dataset describes response in the length of teeth of 10 guinea pigs
treated with Vitamin C. There are six groups produced for three dose levels
(0.5, 1 and 2 mg) and two supplement delivery methods (orange juice and ascorbic
acid).  

There were six groups of the animals: 10 guinea pigs at each
of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery
methods (orange juice or ascorbic acid).

# Confidence intervals

Intuitive interpretation of the data on the chart is that supplement does
affects tooth length, but delivery type efficiency depends on dose. Let's
check this intuitive interpretations by calculating confidence intervals.

## Length per dose

Visually we see that teeth length significantly higher for each level of
supplement dose. This is our assumption. Let's calculate confidence intervals
to see if it is indeed so,
and intuitive comprehension of the chart doesn't trick us.


It's not completely clear if 10 guinea pigs described
in the help page for ToothGrowth dataset are the same animals in each of groups,
or different ones. Even if they are the same, there's no clear indication in the
dataset of particular rodents, so we are going to treat them as if they were
different, and data points aren't paired. If we will still see that changes
are signigicant, then it will be true even if pigs were the same and data points
paired, because paired data is less variable, and less confidence intervals are
narrower.


We calculate confidence intervals for tooth length change for four pair of groups:
dose change from 0.5 to 1 and dose  change from 1 to 2 - for each type of
supplement.
Notice that t.test()'s default value for paired is FALSE, and this is what we want.
We also presume that variability in groups isn't equal and
this is also the default setting for the t test.

```{r}
CheckIntervalByDose <- function(before,after,supp) {
      if(missing(supp)) suppC <- TRUE else suppC <- ToothGrowth$supp == supp
      t.test(ToothGrowth$len[ToothGrowth$dose == after & suppC],
       ToothGrowth$len[ToothGrowth$dose == before & suppC]
       )$conf.int
}

CheckIntervalByDose(0.5, 1, "OJ")
CheckIntervalByDose(1,   2, "OJ")
CheckIntervalByDose(0.5, 1, "VC")
CheckIntervalByDose(1,   2, "VC")
```

We see that the confidence intervals lie entirely above zero in every case,
so it's clear that increasing the dosage of the supplement does indeed increase
tooth length in guinea pigs with 95% confidence level.

Now, it probably wouldn't make much sense for a real research to lose
information we have about the experiment and ignore supplement delivery type,
but just for the training purposes, let's see if the result changes when we do
that.
```{r}
CheckIntervalByDose(0.5, 1)
CheckIntervalByDose(1,   2)
```
Intervals are still entirely above zero, and the distances aren't dramatically
different than before - it's probably because increased variability was
compensated by having twice as much data points, and having more points
increases confidence.

## Result
Assuming data is normally distributed, and using t distribution for safer
approximation for small given number of observations, the confidence interval
show that that change in teeth length is statistically significant for each
level of dose increase, both with separate analysis of two different delivery
ways, and mixed as well. Given this, we can say that vitamin C supplements
do indeed increase tooth length in guinea pigs.